---
title: "time series forecast peak shaving"
author: "Gabrielle Nyirjesy"
date: '2022-12-03'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r}
library(dplyr)
library(xts)
library(lubridate)
library(forecast)
library(ggplot2)
library(Metrics)
library(stringr)
```

# Data pre-processing

```{r}
df <- read.csv('./data/demand_unpivoted.csv')
head(df)
```
```{r}
df$timestamp <- strptime(df$timestamp, format="%Y-%m-%d %H:%M:%S")
head(df)
```

```{r}
df_2018 <- subset(df, df$Date < as.POSIXct('2019-01-01 00:00:00') )
df_2019 <- subset(df, df$Date >= as.POSIXct('2019-01-01 00:00:00') )
df_ts <- data.frame(interval = seq(ymd_hms('2018-01-01 00:00:00'), 
                                by = '15 min',length.out=(length(df$demand))), 
                 data = df$demand)
full_ts <- xts(df_ts$data, order.by=df_ts$interval)

df_ts_train <- data.frame(interval = seq(ymd_hms('2018-01-01 00:00:00'), 
                                by = '15 min',length.out=(length(df_2018$demand))), 
                 data = df_2018$demand)
ts_2018 <- xts(df_ts_train$data, order.by=df_ts_train$interval)

df_ts_test <- data.frame(interval = seq(ymd_hms('2019-01-01 00:00:00'), 
                                by = '15 min',length.out=(length(df_2019$demand))), 
                 data = df_2019$demand)
ts_2019 <- xts(df_ts_test$data, order.by=df_ts_test$interval)

```

# Modeling

## Naive Model

```{r}
model <- naive(ts(df_ts_train$data, frequency = 96), level=c(80,95), h=length(df_ts_test$data))
rmse_error <- sqrt( mean( (df_ts_test$data-model$mean)^2 , na.rm = TRUE ) )
rmse_pct <- round(rmse_error/mean(df_ts_test$data)*100,2)
mean_abs_error <- sum(abs(df_ts_test$data-model$mean), na.rm = TRUE)/length(df_ts_test$data)
mae_pct <- round(mean_abs_error/mean(df_ts_test$data)*100,2)
plot(df_ts$interval, df_ts$data, xlab='Period',ylab='Demand', type='l', main = 'Naive Model: Actual and Predicted Demand',
     sub = paste0('*RMSE: ',round(rmse_error,2), '(',rmse_pct,'%)', ', MAE: ', round(mean_abs_error,2), '(',mae_pct,'%)'))
lines(df_ts_test$interval, model$mean, col='blue')
legend(x='bottomright', legend=c('Actual','Predicted'), col=c('black','blue'),lty=1)
```

## ARIMA Model

```{r}
model <- auto.arima(ts(df_ts_train$data, frequency = 96),
                          stepwise = FALSE,
                          approximation = FALSE,
                          seasonal = FALSE,
                          ic = 'aicc')
model_preds <- forecast(model, level=c(80,95), h=length(df_ts_test$data))
rmse_error <- sqrt( mean( (df_ts_test$data-model_preds$mean)^2 , na.rm = TRUE ) )
rmse_pct <- round(rmse_error/mean(df_ts_test$data)*100,2)
mean_abs_error <- sum(abs(df_ts_test$data-model_preds$mean), na.rm = TRUE)/length(df_ts_test$data)
mae_pct <- round(mean_abs_error/mean(df_ts_test$data)*100,2)
plot(df_ts$interval, df_ts$data, xlab='Period',ylab='Demand', type='l', main = paste0(model_preds$method,' Model: Actual and Predicted Demand'),
     sub = paste0('*RMSE: ',round(rmse_error,2), '(',rmse_pct,'%)', ', MAE: ', round(mean_abs_error,2), '(',mae_pct,'%)'))
lines(df_ts_test$interval, model_preds$mean, col='blue')
legend(x='bottomright', legend=c('Actual','Predicted'), col=c('black','blue'),lty=1)
```

## ETS Model

```{r}
model <- ets(ts(df_ts_train$data, frequency = 96))
model_preds <- forecast(model, level=c(80,95), h=length(df_ts_test$data))
rmse_error <- sqrt( mean( (df_ts_test$data-model_preds$mean)^2 , na.rm = TRUE ) )
rmse_pct <- round(rmse_error/mean(df_ts_test$data)*100,2)
mean_abs_error <- sum(abs(df_ts_test$data-model_preds$mean), na.rm = TRUE)/length(df_ts_test$data)
mae_pct <- round(mean_abs_error/mean(df_ts_test$data)*100,2)
plot(df_ts$interval, df_ts$data, xlab='Period',ylab='Demand', type='l', main = 'ETS Model: Actual and Predicted Demand',
     sub = paste0('*RMSE: ',round(rmse_error,2), '(',rmse_pct,'%)', ', MAE: ', round(mean_abs_error,2), '(',mae_pct,'%)'))
lines(df_ts_test$interval, model_preds$mean, col='blue')
legend(x='bottomright', legend=c('Actual','Predicted'), col=c('black','blue'),lty=1)
```

## STLF Model

```{r}
freq = 96*30.25
model <- stlf(ts(df_ts_train$data, frequency = freq), h=length(df_ts_test$data))
rmse_error <- sqrt( mean( (df_ts_test$data-model$mean)^2 , na.rm = TRUE ) )
rmse_pct <- round(rmse_error/mean(df_ts_test$data)*100,2)
mean_abs_error <- sum(abs(df_ts_test$data-model$mean), na.rm = TRUE)/length(df_ts_test$data)
mae_pct <- round(mean_abs_error/mean(df_ts_test$data)*100,2)
plot(df_ts$interval, df_ts$data, xlab='Period',ylab='Demand', type='l', main = 'STL + ETS(A,N,N) Model: Actual and Predicted Demand',
     sub = paste0('*RMSE: ',round(rmse_error,2), '(',rmse_pct,'%)', ', MAE: ', round(mean_abs_error,2), '(',mae_pct,'%)'))
lines(df_ts_test$interval, model$mean, col='blue')
legend(x='bottomright', legend=c('Actual','Predicted'), col=c('black','blue'),lty=1)
```

```{r}
#Save the predicted demand
write.csv(model$fitted, './data/stlf_predicted_2019_demand.csv')

```

STLF model captures the trend the best, so use this moving forward

# Pipeline for Optimization Model

Train model on 2018 only data, create forecast, update each month and re-train the forecast
Try data smoothing

```{r}
lower_quantile <- quantile(subset(df, df$Date < as.POSIXct('2019-01-01 00:00:00'))$demand, .25)
upper_quantile <- quantile(subset(df, df$Date < as.POSIXct('2019-01-01 00:00:00'))$demand, .75)
inter_quartile_range <- IQR(subset(df, df$Date < as.POSIXct('2019-01-01 00:00:00'))$demand)
lower_outlier_cutoff <- lower_quantile - 1.5*inter_quartile_range
upper_outlier_cutoff <- upper_quantile + 1.5*inter_quartile_range
#Replace outliers in demand
df$smoothed <- ifelse(df$demand < lower_outlier_cutoff, lower_outlier_cutoff, 
                      ifelse(df$demand > upper_outlier_cutoff, upper_outlier_cutoff, df$demand))
plot(df$demand, type='l')
lines(df$smoothed, type='l', col='blue')
```

Create empty column to fill with forecasts

```{r}
df['forecast'] <- NA
start_index = 1
head(df)
```

Forecast for each month and then update the train data with that month's data
for the next month's forecast.

```{r}
res <- c()
for(i in 1:12){
  if(i == 12){
    start_test_period = '2019-12-01 00:00:00'
    end_test_period = '2020-01-01 00:00:00'
  }
  else{
    start_test_period = paste0('2019-',str_pad(i, 2, pad = "0"),'-01 00:00:00')
    end_test_period = paste0('2019-',str_pad(i+1, 2, pad = "0"),'-01 00:00:00')
  }
  # df$log_demand <- log(df$demand)
  df_train <- subset(df, df$Date < as.POSIXct(start_test_period) )
  df_test <- subset(df, (df$Date >= as.POSIXct(start_test_period) )& (df$Date < as.POSIXct(end_test_period) ))
  df_ts <- data.frame(interval = seq(ymd_hms('2018-01-01 00:00:00'), 
                                  by = '15 min',length.out=(length(df$demand))), 
                   data = df$demand)
  full_ts <- xts(df_ts$data, order.by=df_ts$interval)
  
  df_ts_train <- data.frame(interval = seq(ymd_hms('2018-01-01 00:00:00'), 
                                  by = '15 min',length.out=(length(df_train$demand))), 
                   data = df_train$demand)
  df_ts_test <- data.frame(interval = seq(ymd_hms(start_test_period), 
                                  by = '15 min',length.out=(length(df_test$demand))), 
                   data = df_test$demand)
  
  freq = 96*30.25
  model <- stlf(ts(df_ts_train$data, frequency = freq), h=length(df_ts_test$data))
  rmse_error <- sqrt( mean( (df_ts_test$data-model$mean)^2 , na.rm = TRUE ) )
  rmse_pct <- round(rmse_error/mean(df_ts_test$data)*100,2)
  mean_abs_error <- sum(abs(df_ts_test$data-model$mean), na.rm = TRUE)/length(df_ts_test$data)
  mae_pct <- round(mean_abs_error/mean(df_ts_test$data)*100,2)
  plot(df_ts$interval, df_ts$data, xlab='Period',ylab='Demand', type='l', main = 'STL + ETS(A,N,N) Model: Actual and Predicted Demand',
       sub = paste0('*RMSE: ',round(rmse_error,2), '(',rmse_pct,'%)', ', MAE: ', round(mean_abs_error,2), '(',mae_pct,'%)'))
  lines(df_ts_test$interval, model$mean, col='blue')
  legend(x='bottomright', legend=c('Actual','Predicted'), col=c('black','blue'),lty=1)
  res <- c(res, model$mean)
}
```
```{r}
df_ts['forecast'] <- NA
df_ts[35041:length(df_ts$data),]$forecast <- res
```

```{r}
testing <- df_ts[35041:length(df_ts$data),]
rmse_error <- sqrt( mean( (testing$data-testing$forecast)^2 , na.rm = TRUE ) )
rmse_pct <- round(rmse_error/mean(testing$data)*100,2)
mean_abs_error <- sum(abs(testing$data-testing$forecast), na.rm = TRUE)/length(testing$data)
mae_pct <- round(mean_abs_error/mean(testing$data)*100,2)
plot(df_ts$interval, df_ts$data, col='black', xlab='Period',ylab='Demand', type='l', main = 'STL + ETS(A,N,N) Model: Actual and Predicted Demand',sub = paste0('*RMSE: ',round(rmse_error,2), '(',rmse_pct,'%)', ', MAE: ', round(mean_abs_error,2), '(',mae_pct,'%)'))
lines(df_ts$interval, df_ts$forecast, type='l', col='blue')
legend(x='bottomright', legend=c('Actual','Predicted'), col=c('black','blue'),lty=1)
```
```{r}
# write.csv(df_ts, './data/forecasted_demand.csv')
```

